---
title: "PPOL561"
output: 
  html_document:
    css: style.css
    theme: spacelab
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r,include=F}
knitr::opts_chunk$set(error = F,message = F,comment = F,warning = F)
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```



<center><h1>Accelerated Statistics for Public Policy II</h1></center>
<center>**Spring 2020**</center>
<center>**Georgetown University**</center>

<br>



# Course Outline {.tabset .tabset-fade .tabset-pills}

## Course Schedule

<br>
<br>


```{r,echo=F,message=F,results='asis'}
read_csv('Syllabus/spring-2020-timeline.csv') %>%
  select(-`Presentation Group`) %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size = 22) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '4in') %>% 
  column_spec(4,width = '2in') 
```

<br>
<br>
<br>
<br>


## Syllabus

<br>
<br>

- Access an [**Online Version**](Syllabus/PPOL561-Accelerated-Statistics-for-Public-Policy-02.html)
- [**Download**](Syllabus/PPOL561-Accelerated-Statistics-for-Public-Policy-02.pdf) a `.pdf` version locally. 

<br>
<br>
<br>
<br>

## Readings

<br>

List of [**all required and suggested readings**](Readings/reading-list.html)

<br>
<br>
<br>
<br>


## Installation

<br>
<br>

The following are installation instructions for `R` and `RStudio`.

<br>
<br>

**`R` Software**

To install `R`, download `R` from CRAN via the following:
 
- Windows: https://cran.r-project.org/bin/windows/base/ (Links to an external site.)
- Mac: https://cran.r-project.org/bin/macosx/
 
To install `RStudio`, download from the following (scroll to the bottom):

- https://www.rstudio.com/products/rstudio/download/ (Links to an external site.)


<br>
<br>
<br>
<br>

## Recitation

<br>
<br>

Recitation meets most weeks on Wednesdays from 4:00 to 5:00 in **Old North**. Below is the schedule for when there is a lab. 

```{r,echo=F}
read_csv("Syllabus/lab-schedule-spring2020.csv") %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  # kable(., "latex", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size = 20) %>% 
  column_spec(1,width = '1in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '4in') 
```


<br>
<br>
<br>
<br>


## Contact

<br>

#### [Eric Dunford](www.ericdunford.com) (Professor)

- **Office**: 404 Old North
- **Office Hours**: Tuesdays 1pm to 3pm
- **Email**: [eric.dunford@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### [Madeline Pickens](https://mccourt.georgetown.edu/news/profile/maddie-pickens/) (Teaching Assistant)

- **Office Hours**: by appointment 
- **Email**: [mp1595@georgetown.edu](mailto:mp1595@georgetown.edu)

<br>

#### Class

- **Slack**: class communications will primarily take place through [Slack](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members). Please follow the [**_invite link_**](https://join.slack.com/t/ppol561accele-ilo1311/shared_invite/enQtODk3MTAzNjIxNDMxLWMzYmI3OGRjZGE2MTU1NWViNTA4YzQ1ZDYxNDU4NmI3MzNkYjBkNTYxNTkyZTE2YzM3ZTNmNWM3ZTA5NjIyODg) to be added to the class Slack channel.

<br>
<br>
<br>
<br>



# Lecture Materials

<br>
<br>


---
