---
title: "PPOL561"
output: 
  html_document:
    css: style.css
    theme: spacelab
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r,include=F}
knitr::opts_chunk$set(error = F,message = F,comment = F,warning = F)
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```



<center><h1>Accelerated Statistics for Public Policy II</h1></center>
<center>**Spring 2021**</center>
<center>**Georgetown University**</center>

<br>



# Course Outline {.tabset .tabset-fade .tabset-pills}

## Course Schedule

<br>
<br>


```{r,echo=F,message=F,results='asis'}
read_csv('Syllabus/spring-2021-timeline.csv') %>%
  select(-`Presentation Group`) %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size = 22) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '4in') %>% 
  column_spec(4,width = '2in') 
```


<br>
<br>

---

<br>
<br>


## Syllabus

<br>
<br>

- Access an [**Online Version**](Syllabus/PPOL561-Accelerated-Statistics-for-Public-Policy-02.html)
- [**Download**](Syllabus/PPOL561-Accelerated-Statistics-for-Public-Policy-02.pdf) a `.pdf` version locally. 

<br>
<br>

---

<br>
<br>

## Readings

<br>

List of [**all required and suggested readings**](Readings/reading-list.html)

<br>
<br>

---

<br>
<br>

## Virtual Classroom


<br> 


#### [Link to Virtual Zoom Classroom](https://georgetown.zoom.us/j/92576720151?pwd=Umc1R1lWdTlmeU1kVUlnZkxQVWVldz09) (Tuesdays 9:00am - 11:00am)
- **Meeting Id**: 925 7672 0151
- **Pass code**: 43765920
  
<br>
  
#### [Virtual Office Hours](https://georgetown.zoom.us/j/96444963432) (Wednesdays 9:00am - 11:00am)

  
<br>

#### [Lab Room](https://georgetown.zoom.us/j/95121179987?pwd=SGh3TEwyOTNJUkk5aUZYQU1MY2xuQT09) (Wednesdays 10:00am - 11:00am)

<br>

Recurrent Zoom link can also be found on Canvas. Please contact the professor/TA through Slack if any of the Zoom links break. 

<br>
<br>

---

<br>
<br>

## Recitation

<br>
<br>

Recitation meets virtually most weeks on Wednesdays from 10:00am to 11:00am. Below is the schedule for when there is a lab. 

Like the virtual lectures, recitation will take place on Zoom. The zoom link will be the same as the class meeting link. If the link breaks, the teaching assistant will post a new link through the `#labs` tab on Slack. 

```{r,echo=F}
read_csv("Syllabus/lab-schedule-spring2021.csv") %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  # kable(., "latex", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size = 20) %>% 
  column_spec(1,width = '1in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '4in') 
```


<br>
<br>

---

<br>
<br>

## Presentation

<br>
<br>

The presentation is a 10 minute in-class presentation with slides on a paper related to the material we are discussing. These presentations will be done in teams of two. Each team will be responsible for locating a research paper published in a peer-reviewed journal. The presentation should summarize the substantive and statistical issues addressed in the paper and provide context and a critique. 

Students will be evaluate along 5 categories: preparedness, presentation performance, slides, critiques, and timing. Please see the rubric for specifics. 

<br>

<center>

[**Presentation Rubric**](Assignments/Presentations/presentation_rubric_PPOL561.pdf)

</center>

<br>

The **presentation schedule** for the Spring 2021 semester: 

```{r,echo=FALSE,message=FALSE,warning=FALSE}
read_csv(here::here('Students/presentation_schedule.csv')) %>% 
  kable(., "html", booktabs = T,align = c('c','c','c')) %>%
  kable_styling(latex_options = c("striped", "scale_down","repeat_header"),
                full_width = T,
                font_size = 18,
                row_label_position = 'c') 
```

<br>
<br>

---

<br>
<br>

## Installation

<br>
<br>

The following are installation instructions for `R` and `RStudio`.

<br>
<br>

**`R` Software**

To install `R`, download `R` from CRAN via the following:
 
- Windows: https://cran.r-project.org/bin/windows/base/ (Links to an external site.)
- Mac: https://cran.r-project.org/bin/macosx/
 
To install `RStudio`, download from the following (scroll to the bottom):

- https://www.rstudio.com/products/rstudio/download/ (Links to an external site.)


<br>
<br>

---

<br>
<br>


## Contact

<br>

#### [Eric Dunford](http://ericdunford.com/) (Professor)

- **Office Hours**: Wednesdays 9am to 11am
- **Email**: [eric.dunford@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### Emma Farber (Teaching Assistant)

- **Office Hours**: Wednesdays 10am - 12pm (after lab)
- **Email**: [ejf78@georgetown.edu ](mailto:ejf78@georgetown.edu )

<br>

#### Class

- **Slack**: class communications will primarily take place through [Slack](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members). 

<!-- Please follow the [**_invite link_**](https://join.slack.com/t/ppol561accele-ilo1311/shared_invite/enQtODk3MTAzNjIxNDMxLWMzYmI3OGRjZGE2MTU1NWViNTA4YzQ1ZDYxNDU4NmI3MzNkYjBkNTYxNTkyZTE2YzM3ZTNmNWM3ZTA5NjIyODg) to be added to the class Slack channel. -->

<br>
<br>

---

<br>
<br>

# Lecture Materials

<br>
<br>

### _Week 1_: [Course Introduction & Research Design](Lectures/week_01/week-01_ppol561.html)

<br>

### _Week 2_: [Data Wrangling and Presentation in `R`](Lectures/week_02/week-02_ppol561.html)

- **Table of Contents**
  + [Data Wrangling](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#3)
  + [Reading & Writing Data](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#7)
  + [Data Manipulation](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#11)
  + [Piping](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#38)
  + [Joining](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#45)
  + [Reshaping](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#67)
  + [Visualization](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#81)


<br>

### _Week 3_: [Introduction to Causal Inference](Lectures/week_03/week-03-causal-inference-ppol561.html)

- **Table of Contents**
  + [Causal Inference](http://ericdunford.com/ppol561/Lectures/week_03/week-03-causal-inference-ppol561.html#3)
  + [Structural Causal Models (DAGs)](http://ericdunford.com/ppol561/Lectures/week_03/week-03-causal-inference-ppol561.html#7)
  + [Causal Discovery](http://ericdunford.com/ppol561/Lectures/week_03/week-03-causal-inference-ppol561.html#51)
  + [Potential Outcomes](http://ericdunford.com/ppol561/Lectures/week_03/week-03-causal-inference-ppol561.html#74)

<br>

### _Week 4_: [OLS, Confounders & Simulation](Lectures/week_04/week-04_simulation_ppol561.html)

- **Table of Contents**
  + [Simulations](http://ericdunford.com/ppol561/Lectures/week_04/week-04_simulation_ppol561.html#3)
  + [Precision of Estimates](http://ericdunford.com/ppol561/Lectures/week_04/week-04_simulation_ppol561.html#18)
  + [Confounding](http://ericdunford.com/ppol561/Lectures/week_04/week-04_simulation_ppol561.html#25)

<br>

### _Week 5_: [Panel Data & DID](Lectures/week_05/week05-panel_data_and_did-ppol561.html)
- **Table of Contents**
    + [Panel Data](http://ericdunford.com/ppol561/Lectures/week_05/week05-panel_data_and_did-ppol561.html#3)
    + [Difference-in-Difference](http://ericdunford.com/ppol561/Lectures/week_05/week05-panel_data_and_did-ppol561.html#45)

<br>

### _Week 6_: [Instrumental Variables](Lectures/week_06/week-06-instrumental-variables-ppol561.html)
- **Table of Contents**
    + [Instruments](http://ericdunford.com/ppol561/Lectures/week_06/week-06-instrumental-variables-ppol561.html#3)
    + [Two Stage Least Squares (2SLS)](http://ericdunford.com/ppol561/Lectures/week_06/week-06-instrumental-variables-ppol561.html#17)
    + [Multiple Instruments](http://ericdunford.com/ppol561/Lectures/week_06/week-06-instrumental-variables-ppol561.html#50)
    + [Weak Instruments](http://ericdunford.com/ppol561/Lectures/week_06/week-06-instrumental-variables-ppol561.html#56)
    + [Precision of 2SLS](http://ericdunford.com/ppol561/Lectures/week_06/week-06-instrumental-variables-ppol561.html#64)


<br>


### _Week 7_: [Experiments](Lectures/week_7/week-07-experiments-ppol561.html)

- **Table of Contents**
    + [Randomization & Balance](http://ericdunford.com/ppol561/Lectures/week_7/week-07-experiments-ppol561.html#3)
    + [Compliance & Intent-to-Treat](http://ericdunford.com/ppol561/Lectures/week_7/week-07-experiments-ppol561.html#21)
    + [Assignment as Instrument](http://ericdunford.com/ppol561/Lectures/week_7/week-07-experiments-ppol561.html#42)
    + [Attrition](http://ericdunford.com/ppol561/Lectures/week_7/week-07-experiments-ppol561.html#50)
    + [Natural Experiments](http://ericdunford.com/ppol561/Lectures/week_7/week-07-experiments-ppol561.html#63)
    
<br>



<!-- ### _Week 3_: [OLS Review](Lectures/week_03/week-03_ppol561.html) -->

<!-- - **Table of Contents** -->
<!--   + [Using Simulations](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#3) -->
<!--   + [Multivariate Models](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#19) -->
<!--   + [Omitted Variable Bias](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#26) -->
<!--   + [Precision of Estimates](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#38) -->
<!--   + [Model Specification](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#48) -->
<!--   + [Dummy Variables](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#50) -->
<!--   + [Standard Errors](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#71) -->

<!-- <br>   -->

<!-- ### _Week 4_: [Binary Outcomes](Lectures/week_04/week-04_ppol561.html) -->

<!-- - **Table of Contents** -->
<!--   + [Binary Dependent Variables](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#4) -->
<!--   + [Maximum Likelihood Estimation](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#40) -->
<!--   + [Substantive Effects](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#50) -->
<!--   + [Simulating confidence intervals around substantive effects](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#76) -->


<!-- <br>   -->

<!-- ### _Week 5_: [Ordered and Multinomial Outcomes](Lectures/week_05/week-05_ordered-multinomial_ppol561.html) -->

<!-- - **Table of Contents** -->
<!--   + [Ordered Outcomes](http://ericdunford.com/ppol561/Lectures/week_05/week-05_ordered-multinomial_ppol561.html#3) -->
<!--   + [Multinomial Outcomes](http://ericdunford.com/ppol561/Lectures/week_05/week-05_ordered-multinomial_ppol561.html#31) -->


<!-- <br>   -->

<!-- ### _Week 6_: [Selection](Lectures/week_06/week-06-selection-ppol561.html) -->

<!-- - **Table of Contents** -->
<!--   + [Selection Problems](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#3) -->
<!--   + [The Selection Process](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#13) -->
<!--   + [Exogenous Selection](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#27) -->
<!--   + [Selection on the Dependent Variable](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#32) -->
<!--   + [Incidental Selection](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#36) -->
<!--   + [Heckman Selection Model](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#45) -->
<!--   + [Example: Women's Labor Force Participation](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#62) -->

<!-- <br>   -->

<!-- ### _Week 7_: [Introduction to Causal Inference](Lectures/week_07/week-07-causal-inference-ppol561.html) -->

<!-- - **Table of Contents** -->
<!--     + [Causal Inference](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#3) -->
<!--     + [Structural Causal Models (DAGs)](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#7) -->
<!--     + [Interventions and Adjustments](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#35) -->
<!--     + [Causal Discovery](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#60) -->
<!--     + [Potential Outcomes](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#82) -->

<!-- <br>   -->

<!-- ### _Week 8_: [Panel Data & DID](Lectures/week_08/week08-panel_data_and_did-ppol561.html) -->
<!-- - **Table of Contents** -->
<!--     + [Panel Data](http://ericdunford.com/ppol561/Lectures/week_08/week08-panel_data_and_did-ppol561.html#3) -->
<!--     + [Difference-in-Difference](http://ericdunford.com/ppol561/Lectures/week_08/week08-panel_data_and_did-ppol561.html#45) -->

<!-- <br> -->

<!-- ### _Week 9_: [Instrumental Variables](Lectures/week_09/week-09-instrumental-variables-ppol561.html) -->
<!-- - **Table of Contents** -->
<!--     + [Instruments](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#3) -->
<!--     + [Two Stage Least Squares (2SLS)](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#16) -->
<!--     + [Multiple Instruments](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#48) -->
<!--     + [Weak Instruments](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#54) -->
<!--     + [Precision of 2SLS](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#62) -->



<!-- <br>  -->


<!-- ### _Week 11_: [Experiments](Lectures/week_11/week-11-experiments-ppol561.html) -->

<!-- - **Table of Contents** -->
<!--     + [Randomization & Balance](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#3) -->
<!--     + [Compliance & Intent-to-Treat](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#16) -->
<!--     + [Assignment as Instrument](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#35) -->
<!--     + [Attrition](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#42) -->
<!--     + [Natural Experiments](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#54) -->

<!-- <br>  -->

<!-- ### _Week 12_: [Regression Discontinuity](Lectures/week_12/week12-regression-discontinuity-ppol561.html) -->

<!-- - **Table of Contents** -->
<!--     + [Randomization & Balance](http://ericdunford.com/ppol561/Lectures/week_12/week12-regression-discontinuity-ppol561.html#3) -->
<!--     + [More flexible RD models](http://ericdunford.com/ppol561/Lectures/week_12/week12-regression-discontinuity-ppol561.html#30) -->
<!--     + [Limitations & Diagnostics](http://ericdunford.com/ppol561/Lectures/week_12/week12-regression-discontinuity-ppol561.html#49)    -->


<!-- <br>  -->

<!-- ### _Week 13_: [Matching](Lectures/week_13/week13-matching-ppol561.html) -->

<!-- - **Table of Contents** -->
<!--     + [_Refresh_ on Causal Effects](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#3) -->
<!--     + [Model Dependency](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#14) -->
<!--     + [Matching](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#35) -->
<!--     + [Matching Methods](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#70) -->
<!--     + [Simulation Example](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#109) -->

<!-- <br> -->

<!-- ### _Week 14_: [Synthetic Control](Lectures/week_14/week14-synthetic-control-ppol561.html) -->

<!-- - **Table of Contents** -->
<!--     + [Case Comparisons](http://ericdunford.com/ppol561/Lectures/week_14/week14-synthetic-control-ppol561.html#3) -->
<!--     + [Synthetic Control](http://ericdunford.com/ppol561/Lectures/week_14/week14-synthetic-control-ppol561.html#9) -->
<!--     + [Example: _Proposition 99_](http://ericdunford.com/ppol561/Lectures/week_14/week14-synthetic-control-ppol561.html#26) -->
<!--     + [Inference](http://ericdunford.com/ppol561/Lectures/week_14/week14-synthetic-control-ppol561.html#42) -->

  