---
title: "PPOL561"
output: 
  html_document:
    css: style.css
    theme: spacelab
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

```{r,include=F}
knitr::opts_chunk$set(error = F,message = F,comment = F,warning = F)
# Packages for reading in the tables
require(tidyverse)
require(kableExtra)
```



<center><h1>Accelerated Statistics for Public Policy II</h1></center>
<center>**Spring 2020**</center>
<center>**Georgetown University**</center>

<br>



# Course Outline {.tabset .tabset-fade .tabset-pills}

## Course Schedule

<br>
<br>


```{r,echo=F,message=F,results='asis'}
read_csv('Syllabus/spring-2020-timeline.csv') %>%
  select(-`Presentation Group`) %>% 
  {.[is.na(.)]="";.} %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size = 22) %>% 
  column_spec(1,width = '.5in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '4in') %>% 
  column_spec(4,width = '2in') 
```


<br>
<br>

---

<br>
<br>


## Instructional Continuity

<br>

#### The Plan 

Due to the COVID-19 outbreak, PPOL561 will meet virtually for the remainder of the semester. We will use a number of third-party tools to help coordinate questions, course materials, and virtual class meetings. The aim is to make the transition from in-person to virtual meetings as seamless as possible. I completely acknowledge that there will be bumps along the way. We'll do our best to work through them as a group. 

I thank you in advance for your patience and understanding as we transition into an online space. Also, I would greatly appreciate feedback regarding what is and is not working so that we can work out the kinks together. I encourage you to reach out to me directly via [**Slack**](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members) or email if you have any suggestions and/or are running into any issues. If you'd rather comment _anonymously_, I've created a space to do so via the following [**Google Form**](https://forms.gle/LLsBWxuvWCe68yfZ6).  

Outlined below is the approach we'll take to continuing instruction online. Specifically, I outline the **tools** we'll be using and how we'll be using them for **_virtual lecture_**, **_office hours_**, and **_communication_**. Finally, I outline how student requirements for the course (**_homework_**, **_presentations_**, and **_exams_**) will operate. 

<u>Please note that this plan is subject to change given feedback/experience</u>. Any and all changes to the plan will be posted on the course website and Slack via the `#general` channel, so please check both regularly for any updates.

Finally, at your earliest convenience, please complete the [**"Transitioning to Online Instruction" survey**](https://forms.gle/xij1wGLCkhqVb2ji9). This survey will help me get a better sense of where everyone is at with respect to resources and personal situations.


<br>

#### Tools

The following is a summary of the tools we'll be using. You'll note that some (most) we've already integrated into the course. Please make sure that you are able to utilize all these resources on your personal machine prior to the start of class on Wednesday, March 18. Note that many of these resources require a dedicated laptop/desktop with a web-camera and microphone, rather than a tablet or phone. In addition, it is important that you also have a reliable internet connection. Please let me know if you do not have access to a computer or reliable internet. 

<br>

- [**Slack**](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members) - for coordinating communication between class, TA, and professor;
- [**Zoom**](https://zoom.us/download) - web-conferencing platform for virtual lecture and office hours;
- [**Google Calendar**](https://calendar.google.com/calendar/embed?src=georgetown.edu_nm9l16pem9re7adpr066ccfkko%40group.calendar.google.com&ctz=America%2FNew_York) - for coordinating zoom links and meeting times; 
- [**Canvas**](https://georgetown.instructure.com/courses/106244) - for course materials, lecture recordings, and exams.

<br>

#### Virtual Lecture
We will use [**Zoom**](https://zoom.us/download) (a web-conferencing platform) to hold class each week. Class will meet at its regularly scheduled time each week (Wednesdays 12:30 - 3:00). If you do not have Zoom, you can download it [**here**](https://zoom.us/download) prior to the start of class. The **_link to the class zoom web-conference can be found on the [Google Calendar](https://calendar.google.com/calendar/embed?src=georgetown.edu_nm9l16pem9re7adpr066ccfkko%40group.calendar.google.com&ctz=America%2FNew_York)_**. If that link is broken, please refer to the `#general` channel on Slack for information regarding the new link.

When entering lecture, please ensure:

- Your microphone is muted, unless you're asking a question.
- Your video is on (unless I direct you to turn it off due to streaming issues).
- To ask questions via the chat feature or to use the "hands-up" icon.

**_All lectures will be recorded and posted to Canvas for students that are unable to join during the scheduled meeting time._** If you're unable to join in the lecture in person due to discrepancies in time zone or persistent computer issues, please let me know. Attendance will be taken directly from the attendees of the scheduled zoom session.  

<br>

#### Office Hours
I will use [**Zoom**](https://zoom.us/download) to hold office hours each week. Office hours will be conducted in a common space. Students can feel free to jump on or off the call at their own leisure. I will managed the entry of students into the space using the "Waiting Room" feature in Zoom. If you log on and you're kept in the waiting room, please be patient as I finish up with another student (this is the digital equivalent of sitting on the couch outside my office). The **_Zoom web-conference link can be found on the class [Google Calendar](https://calendar.google.com/calendar/embed?src=georgetown.edu_nm9l16pem9re7adpr066ccfkko%40group.calendar.google.com&ctz=America%2FNew_York)_**, else please refer to the `#general` channel on Slack for information regarding the new link.

If students wish to coordinate a one-on-one meeting with the Professor or the TA, please reach out via Slack. We'll find a time that mutually works and schedule a call via a Google Calendar meeting. Please accept that meeting request to confirm the appointment.

<br>

#### Recitation
Like the virtual lectures, recitation will take place on Zoom. The link to the zoom meeting will be posted on the course calendar. If the link breaks, the teaching assistant will post a new link through the `#labs` tab on Slack. 

<br>

#### Communication 
We will continue to use [**Slack**](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members) as our primary communication tool. Please check Slack regularly for updates and announcements. 

I've created a new `#lecture` channel that we'll use to handle all lecture related questions. Specifically, for questions that occur to you after lecture or if you're watching the lecture asynchronously, please use the `#lecture` channel for your questions. If you're watching the lecture live via Zoom, we'll use the chat feature in Zoom. If there are sufficient questions on, I may opt to make a supplementary video answering those

In addition, we have the `#general`, `#issues`, and `#labs` channels for posting questions/solutions for the group. Slack is a powerful tool designed exactly for the circumstances we find ourselves in, so I encourage everyone to utilize it. 

<br>

#### Calendar

All students enrolled in the course will receive an invitation to a class calendar hosted via [**Google Calendar**](https://calendar.google.com/calendar/embed?src=georgetown.edu_nm9l16pem9re7adpr066ccfkko%40group.calendar.google.com&ctz=America%2FNew_York). Each scheduled meeting contains the [**Zoom**](https://zoom.us/download) link to the class' virtual meeting space. If the class link breaks for some reason, please check the `#general` channel on [**Slack**](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members) for an updated link.

<br>

<iframe src="https://calendar.google.com/calendar/b/1/embed?height=400&amp;wkst=1&amp;bgcolor=%23ffffff&amp;ctz=America%2FNew_York&amp;src=Z2VvcmdldG93bi5lZHVfbm05bDE2cGVtOXJlN2FkcHIwNjZjY2Zra29AZ3JvdXAuY2FsZW5kYXIuZ29vZ2xlLmNvbQ&amp;color=%23E67C73&amp;title=Virtual%20Meeting%20Calendar%20(PPOL561)&amp;showTitle=0&amp;showNav=1&amp;showDate=1&amp;showPrint=0&amp;showTabs=0&amp;showCalendars=0&amp;showTz=1&amp;mode=AGENDA" style="border-width:0" width="800" height="600" frameborder="0" scrolling="no"></iframe>

<br>

#### Class Presentations

Student teams are required to make in-class presentations in PPOL561 at the start of each class (see the "Presentations" tab). These 10-minute presentations will continue as planned. One student from the presenting team will share their screen containing the slides and the team will give their presentation digitally in Zoom (all others should mute their microphone). After they've given their presentation, we'll leave it open for Q&A. Student teams should coordinate via Zoom to prepare their materials. If a team runs into issues in preparing their presentations, please reach out to the Professor.

<br>

#### Homework

Assignments will continue to be distributed and collected on Canvas. Please continue to render `.Rmd` output as a `.pdf` to make it easier for the Teaching Assistant to download and grade the assignments. Grades and comments will be posted on Canvas.

<br>

#### Exams

We are scheduled to take two exams in this course (the midterm on March 25 and the final on May 8). The exams will continue as scheduled and will be conducted on Canvas. Given concerns over requisite hardware and internet reliability, we will no longer be using Protorio to conduct the exam. Rather, the exam will be posted on Canvas where students will be able access it at the time the exam is scheduled to start in the syllabus. Students must submit the exam within 2 hours of the start time. The exams are closed book and student must adhere to the [academic integrity standards](https://georgetown.app.box.com/s/bibfmpo93061uxmwir29) set by the university. 

Materials to prepare for the exam will be distributed to the class before the scheduled exam date (March 18 for the Midterm; April 22 for the Final).


<br>
<br>

---

<br>
<br>


## Syllabus

<br>
<br>

- Access an [**Online Version**](Syllabus/PPOL561-Accelerated-Statistics-for-Public-Policy-02.html)
- [**Download**](Syllabus/PPOL561-Accelerated-Statistics-for-Public-Policy-02.pdf) a `.pdf` version locally. 

<br>
<br>

---

<br>
<br>

## Readings

<br>

List of [**all required and suggested readings**](Readings/reading-list.html)

<br>
<br>

---

<br>
<br>

## Presentation

<br>
<br>

The presentation is a 10 minute in-class presentation with slides on a paper related to the material we are discussing. These presentations will be done in teams of two. Each team will be responsible for locating a research paper published in a peer-reviewed journal. The presentation should summarize the substantive and statistical issues addressed in the paper and provide context and a critique. 

Students will be evaluate along 5 categories: preparedness, presentation performance, slides, critiques, and timing. Please see the rubric for specifics. 

<br>

<center>

[**Presentation Rubric**](Assignments/Presentations/presentation_rubric_PPOL561.pdf)

</center>

<br>

The **presentation schedule** for the Spring 2020 semester: 

```{r,echo=FALSE,message=FALSE,warning=FALSE}
read_csv(here::here('Students/presentation_schedule.csv')) %>% 
  mutate(Week=c(3:9,11:14)) %>% 
  select(Group,Week,everything()) %>% 
  {.[is.na(.)]="";.} %>% 
  select(-contains('email')) %>% 
  kable(., "html", booktabs = T,align = c('c','c','c')) %>%
  kable_styling(latex_options = c("striped", "scale_down","repeat_header"),
                full_width = T,
                font_size = 18,
                row_label_position = 'c') 
```

<br>
<br>

---

<br>
<br>

## Installation

<br>
<br>

The following are installation instructions for `R` and `RStudio`.

<br>
<br>

**`R` Software**

To install `R`, download `R` from CRAN via the following:
 
- Windows: https://cran.r-project.org/bin/windows/base/ (Links to an external site.)
- Mac: https://cran.r-project.org/bin/macosx/
 
To install `RStudio`, download from the following (scroll to the bottom):

- https://www.rstudio.com/products/rstudio/download/ (Links to an external site.)


<br>
<br>

---

<br>
<br>

## Recitation

<br>
<br>

Recitation meets most weeks on Wednesdays from 4:00 to 5:00 in **Old North G-02**. Below is the schedule for when there is a lab. 

```{r,echo=F}
read_csv("Syllabus/lab-schedule-spring2020.csv") %>% 
  kable(., "html", booktabs = T,align = c('c','l','l')) %>%
  # kable(., "latex", booktabs = T,align = c('c','l','l')) %>%
  kable_styling(latex_options = c("striped", 
                                  "scale_down","repeat_header"),
                full_width = T,font_size = 20) %>% 
  column_spec(1,width = '1in') %>% 
  column_spec(2,width = '1in') %>% 
  column_spec(3,width = '4in') 
```


<br>
<br>

---

<br>
<br>


## Contact

<br>

#### [Eric Dunford](http://ericdunford.com/) (Professor)

- **Office**: 404 Old North
- **Office Hours**: Tuesdays 1pm to 3pm
- **Email**: [eric.dunford@georgetown.edu](mailto:eric.dunford@georgetown.edu)

<br>

#### [Madeline Pickens](https://mccourt.georgetown.edu/news/profile/maddie-pickens/) (Teaching Assistant)

- **Office Hours**: by appointment 
- **Email**: [mp1595@georgetown.edu](mailto:mp1595@georgetown.edu)

<br>

#### Class

- **Slack**: class communications will primarily take place through [Slack](https://get.slack.help/hc/en-us/articles/218080037-Getting-started-for-new-members). 

<!-- Please follow the [**_invite link_**](https://join.slack.com/t/ppol561accele-ilo1311/shared_invite/enQtODk3MTAzNjIxNDMxLWMzYmI3OGRjZGE2MTU1NWViNTA4YzQ1ZDYxNDU4NmI3MzNkYjBkNTYxNTkyZTE2YzM3ZTNmNWM3ZTA5NjIyODg) to be added to the class Slack channel. -->

<br>
<br>

---

<br>
<br>

# Lecture Materials

<br>
<br>

### _Week 1_: [Course Introduction & Research Design](Lectures/week_01/week-01_ppol561.html)

<br>

### _Week 2_: [Data Wrangling and Presentation in `R`](Lectures/week_02/week-02_ppol561.html)

- **Table of Contents**
  + [Data Wrangling](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#3)
  + [Reading & Writing Data](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#7)
  + [Data Manipulation](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#11)
  + [Piping](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#38)
  + [Joining](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#45)
  + [Reshaping](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#67)
  + [Visualization](http://ericdunford.com/ppol561/Lectures/week_02/week-02_ppol561.html#82)

- **Walkthrough**:
  + Generating regression tables using `stargazer` - [`.rmd`](Lectures/week_02/practice/week02-Walkthrough-01_Tables.Rmd), [`.html`](Lectures/week_02/practice/week02-Walkthrough-01_Tables.html)

- **Practice**:
  + Practice  <a href="Lectures/week_02/practice/week02-practice-01_Data-Wrangling.Rmd" download> wrangling data (`.Rmd`)</a> then compare your results to the <a href="Lectures/week_02/practice/week02-practice-01_Data-Wrangling_ANSWERS.Rmd" download> answer key (`.Rmd`)</a>
  + Practice <a href="Lectures/week_02/practice/week02-practice-02_joining-reshaping.Rmd" download> joining and reshaping data (`.Rmd`)</a> then compare your results to the <a href="Lectures/week_02/practice/week02-practice-02_joining-reshaping_ANSWERS.Rmd" download> answer key (`.Rmd`)</a>
  + Practice <a href="Lectures/week_02/practice/week02-practice-03_Data-Vizualization.Rmd" download> data visualization using `ggplot2` (`.Rmd`)</a> then compare your results to the <a href="Lectures/week_02/practice/week02-practice-03_Data-Vizualization_ANSWERS.Rmd" download> answer key (`.Rmd`)</a>


<br>

### _Week 3_: [OLS Review](Lectures/week_03/week-03_ppol561.html)

- **Table of Contents**
  + [Using Simulations](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#3)
  + [Multivariate Models](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#19)
  + [Omitted Variable Bias](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#26)
  + [Precision of Estimates](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#38)
  + [Model Specification](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#48)
  + [Dummy Variables](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#50)
  + [Standard Errors](http://ericdunford.com/ppol561/Lectures/week_03/week-03_ppol561.html#71)
  
<br>  
  
### _Week 4_: [Binary Outcomes](Lectures/week_04/week-04_ppol561.html)

- **Table of Contents**
  + [Binary Dependent Variables](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#4)
  + [Maximum Likelihood Estimation](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#40)
  + [Substantive Effects](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#50)
  + [Simulating confidence intervals around substantive effects](http://ericdunford.com/ppol561/Lectures/week_04/week-04_ppol561.html#76)
  
  
<br>  
  
### _Week 5_: [Ordered and Multinomial Outcomes](Lectures/week_05/week-05_ordered-multinomial_ppol561.html)

- **Table of Contents**
  + [Ordered Outcomes](http://ericdunford.com/ppol561/Lectures/week_05/week-05_ordered-multinomial_ppol561.html#3)
  + [Multinomial Outcomes](http://ericdunford.com/ppol561/Lectures/week_05/week-05_ordered-multinomial_ppol561.html#31)
  
  
<br>  
  
### _Week 6_: [Selection](Lectures/week_06/week-06-selection-ppol561.html)

- **Table of Contents**
  + [Selection Problems](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#3)
  + [The Selection Process](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#13)
  + [Exogenous Selection](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#27)
  + [Selection on the Dependent Variable](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#32)
  + [Incidental Selection](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#36)
  + [Heckman Selection Model](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#45)
  + [Example: Women's Labor Force Participation](http://ericdunford.com/ppol561/Lectures/week_06/week-06-selection-ppol561.html#62)
  
<br>  
  
### _Week 7_: [Introduction to Causal Inference](Lectures/week_07/week-07-causal-inference-ppol561.html)

- **Table of Contents**
    + [Causal Inference](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#3)
    + [Structural Causal Models (DAGs)](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#7)
    + [Interventions and Adjustments](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#35)
    + [Causal Discovery](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#60)
    + [Potential Outcomes](http://ericdunford.com/ppol561/Lectures/week_07/week-07-causal-inference-ppol561.html#82)
  
<br>  
  
### _Week 8_: [Panel Data & DID](Lectures/week_08/week08-panel_data_and_did-ppol561.html)
- **Table of Contents**
    + [Panel Data](http://ericdunford.com/ppol561/Lectures/week_08/week08-panel_data_and_did-ppol561.html#3)
    + [Difference-in-Difference](http://ericdunford.com/ppol561/Lectures/week_08/week08-panel_data_and_did-ppol561.html#45)
   
<br>

### _Week 9_: [Instrumental Variables](Lectures/week_09/week-09-instrumental-variables-ppol561.html)
- **Table of Contents**
    + [Instruments](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#3)
    + [Two Stage Least Squares (2SLS)](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#16)
    + [Multiple Instruments](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#48)
    + [Weak Instruments](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#54)
    + [Precision of 2SLS](http://ericdunford.com/ppol561/Lectures/week_09/week-09-instrumental-variables-ppol561.html#62)

    
    
<br> 


### _Week 11_: [Experiments](Lectures/week_11/week-11-experiments-ppol561.html)

- **Table of Contents**
    + [Randomization & Balance](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#3)
    + [Compliance & Intent-to-Treat](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#16)
    + [Assignment as Instrument](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#35)
    + [Attrition](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#42)
    + [Natural Experiments](http://ericdunford.com/ppol561/Lectures/week_11/week-11-experiments-ppol561.html#54)
    
<br> 
    
### _Week 12_: [Regression Discontinuity](Lectures/week_12/week12-regression-discontinuity-ppol561.html)

- **Table of Contents**
    + [Randomization & Balance](http://ericdunford.com/ppol561/Lectures/week_12/week12-regression-discontinuity-ppol561.html#3)
    + [More flexible RD models](http://ericdunford.com/ppol561/Lectures/week_12/week12-regression-discontinuity-ppol561.html#30)
    + [Limitations & Diagnostics](http://ericdunford.com/ppol561/Lectures/week_12/week12-regression-discontinuity-ppol561.html#49)   
    
    
<br> 
    
### _Week 13_: [Matching](Lectures/week_13/week13-matching-ppol561.html)

- **Table of Contents**
    + [_Refresh_ on Causal Effects](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#3)
    + [Model Dependency](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#14)
    + [Matching](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#35)
    + [Matching Methods](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#70)
    + [Simulation Example](http://ericdunford.com/ppol561/Lectures/week_13/week13-matching-ppol561.html#109)

    
    
  