---
title: 
    <font class = "title-panel"> PPOL561 | Accelerated Statistics for Public Policy II</font>
  <font size=6, face="bold"> Week 13 </font> 
  <br>
  <br>
  <font size=100, face="bold"> Ordered and Multinomial Outcomes </font>
author: 
  <font class = "title-footer"> 
  &emsp;Prof. Eric Dunford &emsp;&#9670;&emsp; Georgetown University &emsp;&#9670;&emsp; McCourt School of Public Policy &emsp;&#9670;&emsp; eric.dunford@georgetown.edu</font>
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: "gu-theme.css"
    nature:
      highlightStyle: github
      beforeInit: "macros.js"
      countIncrementalSlides: False
      highlightLines: true
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T,message=F,error=F,warning = F)
require(tidyverse)
```

layout: true

<div class="slide-footer"><span> 
PPOL561 | Accelerated Statistics for Public Policy II

&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;

Week 13 <!-- Week of the Footer Here -->

&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;&emsp;

Ordered and Multinomial Outcomes <!-- Title of the lecture here -->

</span></div> 

---
class: outline

# Outline for Today 

<br><br>
- Modeling **ordered outcomes**

<br><br>
- Modeling **nominal outcomes**


---

class: newsection

# Ordered Outcomes

---

### Ordered outcomes

- Ordered dependent variables are outcomes that retain some inherent but discrete ordering. 

  - **Likert Scales**: "Strongly Disagree", "Disagree", "Somewhat Disagree", "Somewhat Agree", ...
  - **Policy options**: "privatize social security", "paritally privatize", "leave unchanged"
  - **Ranks**: "some high School", "high school grad", "some college", "college grad", etc.

- The distance between each category is of different unknown sizes. 

- When the proper ordering of a variable is ambiguous, multinomial models should be considered.

---

### Why not OLS?

<br>

- Encounter the same problems as when using regression with a binary outcome:
  + _heteroskedastic_
  + _non-sensical predictions_
  
- Requires us to assume that the **distance between categories is equal**.

- Difficult to interpret

---

### Latent Variable Approach 
<br><br>
Recall when discussing binary models we spoke about things in terms of a **latent variable** $y^*$. 

$$y_i^* = \beta_0 + \beta_1x_i + \epsilon_i$$

When $y^*$ is above a threshold $\tau_1$ we observe $y = 1$, otherwise we observe $y = 0$. 

In this model, we assumed that the cutoff was 0 ( $\tau_1 = 0$ ). 

---

```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
set.seed(123)
tibble(x = rnorm(500)) %>% 
  mutate(ystar = 1*x + rnorm(500),
         color = ifelse(ystar>0,"steelblue","grey30")) %>% 
  ggplot(aes(x,ystar,color=color)) +
  geom_point(size=3,alpha=.7) +
  geom_hline(yintercept = 0,color="darkred",size=1,alpha=.5) +
  geom_label(inherit.aes = F,x=-2.5,y=3,label="y = 1",
            size=8,color="steelblue") +
  geom_label(inherit.aes = F,x=-2.5,y=-3,label="y = 0",
            size=8,color="grey30") +
  geom_label(inherit.aes = F,x=0,y=0,label="cutoff",
             size=8,color="darkred") +
  scale_color_identity() +
  xlim(-3,3) +
  labs(y=latex2exp::TeX("$y^*$"),
       title="Latent Variable for a Binary Response") +
  theme_bw() +
  theme(axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 20),
        plot.title = element_text(family='serif',face='bold',size = 20)) 
```


---

```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
set.seed(123)
tibble(x = rnorm(500)) %>% 
  mutate(ystar = 1*x + rnorm(500),
         color = "grey30",
         color = ifelse(ystar>=-1.5 & ystar < 1.5,"steelblue",color),
         color = ifelse(ystar>=1.5,"orange",color)
         ) %>% 
  ggplot(aes(x,ystar,color=color)) +
  geom_point(size=3,alpha=.7) +
  
  
  # Third label
    geom_label(inherit.aes = F,x=-2.5,y=2.5,label="y = 3",
            size=8,color="orange") +
 
  
  # Second cut point
  geom_hline(yintercept = 1.5,color="darkred",size=1,alpha=.5) +
  geom_label(inherit.aes = F,x=-2.5,y=0,label="y = 2",
            size=8,color="steelblue") +
  geom_label(inherit.aes = F,x=0,y=1.5,label="cutoff 2",
             size=8,color="darkred") +
  
  
  # First cut point
  geom_hline(yintercept = -1.5,color="darkred",size=1,alpha=.5) +
  geom_label(inherit.aes = F,x=-2.5,y=-2.5,label="y = 1",
            size=8,color="grey30") +
  geom_label(inherit.aes = F,x=0,y=-1.5,label="cutoff 1",
             size=8,color="darkred") +
  
  scale_color_identity() +
  xlim(-3,3) +
  labs(y=latex2exp::TeX("$y^*$"),
       title="Latent Variable for an Ordered Response") +
  theme_bw() +
  theme(axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 20),
        plot.title = element_text(family='serif',face='bold',size = 20)) 
```

---


```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
set.seed(123)
tibble(x = rnorm(500)) %>% 
  mutate(ystar = 1*x + rnorm(500),
         color = "grey30",
         color = ifelse(ystar>=-1.5 & ystar < 1.5,"steelblue",color),
         color = ifelse(ystar>=1.5,"orange",color),
         ystar2 = 1,
         ystar2 = ifelse(ystar>=-1.5 & ystar2 < 1.5,2,ystar2),
         ystar2 = ifelse(ystar>=1.5,3,ystar2)
         ) %>% 
  ggplot(aes(x,factor(ystar2),color=color)) +
  geom_point(size=3,alpha=.7) +

  
  scale_color_identity() +
  xlim(-3,3) +
  labs(y="y",
       title="Ordered Response") +
  theme_bw() +
  theme(axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 20),
        plot.title = element_text(family='serif',face='bold',size = 20)) 
```


---

The probability of observing a specific response in the ordering _depends on which cut points
$y^∗$ falls in-between_.

- when <font color = "grey"> $y^* \le \tau_1$ </font>, we observe <font color = "grey"> $y=1$ </font>
- when <font color = "steelblue"> $\tau_1 \le y^* < \tau_2$ </font>, we observe <font color = "steelblue"> $y=2$ </font>
- when <font color = "orange"> $\tau_2 \ge y^*$ </font>, we observe <font color = "orange"> $y=3$ </font>

```{r,echo=F,fig.align="center",fig.width=9,fig.height=4.5,cache=T,dpi=300}
set.seed(123)
tibble(x=seq(-3,3,.1),pr=dnorm(seq(-3,3,.1))) %>% 
  mutate(pr1 = ifelse(x <= -1,pr,0),
         x1 = ifelse(x <= - 1,x,-1),
         pr2 = ifelse(x >= -1 & x <= 1 ,pr,0.2419707),
         x2 = ifelse(x >= -1 & x <= 1,x,1),
         pr3 = ifelse(x >= 1 ,pr,0),
         x3 = ifelse(x> 1,x,1),
         ) %>% 
ggplot(aes(x,pr)) +
  geom_area(alpha=.5,fill="grey") +
  geom_area(aes(y=pr1,x1),alpha=.5,fill="grey30") +
  geom_area(aes(y=pr2,x2),alpha=.5,fill="steelblue") +
  geom_area(aes(y=pr3,x3),alpha=.5,fill="orange") +
  geom_vline(xintercept = -1,color="darkred",size=1) +
  geom_vline(xintercept = 1,color="darkred",size=1) +
  geom_label(x=-1,y=.4,label=latex2exp::TeX("$\\tau_1"),size=6,color="darkred") +
  geom_label(x=1,y=.4,label=latex2exp::TeX("$\\tau_2"),size=6,color="darkred") +
  geom_text(x=-2,y=.42,label="y = 1",size=6,color="grey30") +
  geom_text(x=0,y=.42,label="y = 2",size=6,color="steelblue") +
  geom_text(x=2,y=.42,label="y = 3",size=6,color="orange") +
  ylim(0,.45) +
  labs(x = latex2exp::TeX("$\\y^*$"),
       y="Density",
       title="") +
  theme_bw() +
  theme(plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 20)) 
```


---

### Identification
<br><br>
- To identify the model, we need to do one of two things

  - set the **intercept** ( $\beta_0$ ) **equal to 0**
  
  - set the ** $\tau_0$ ** (our $-\infty$ cut point) ** equal to 0**
  
- The `polr()` function from the `MASS` package (which we'll use to estimate a ordered probit and logit model) opts to set the intercept to 0 by default

---

### Parallel Regressions Assumption

<br><br>

- When running the an ordered logit or probit model, we inherently make an assumption about "**parallel regressions**".

- The effect of the coefficients is the same across all cutpoint ranges, i.e. the effect of $x$ on $y$ does not change when we move from one threshold to the next.

- Because the relationship between all pairs of groups is the same, there is **_only one set of coefficients_** (only one model).  If this was not the case, we would need different models to describe the relationship between each pair of outcome groups. 

---

```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
x = seq(-5,5,.1)
y1 = pnorm(x-2)
y2 = pnorm(x)
y3 = pnorm(x+2)
bind_rows(tibble(x,y=y1,color="orange"),
          tibble(x,y=y2,color="steelblue"),
          tibble(x,y=y3,color="grey30")) %>% 
  mutate(color = factor(color,levels = c("grey30","steelblue","orange"))) %>% 
ggplot(aes(x,y,color=color)) +
  geom_line(size=2) +
  scale_color_identity() +
  labs(y ="Probability",
       title="Parallel Regressions") +
  # xlim(-2,2) +
  theme_bw() +
  theme(plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 20)) 
```

---

```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
x = seq(-5,5,.1)
y1 = pnorm( 5*(x-2) )
y2 = pnorm(2*x)
y3 = pnorm( .8*(x+2) )
bind_rows(tibble(x,y=y1,color="orange"),
          tibble(x,y=y2,color="steelblue"),
          tibble(x,y=y3,color="grey30")) %>% 
  mutate(color = factor(color,levels = c("grey30","steelblue","orange"))) %>% 
ggplot(aes(x,y,color=color)) +
  geom_line(size=2) +
  scale_color_identity() +
  labs(y ="Probability",
       title="Not Parallel Regressions") +
  # xlim(-2,2) +
  theme_bw() +
  theme(plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 20)) 
```


---

### Predicted Probabilities

The key to calculating the predicted probabilities of any ordered outcome is **_keeping track of where the cutpoints fall_**. 

<br>

.center[
|               |                                |
| --------------|--------------------------------|
| $pr(y_i = 1)$ | $\Phi(\tau_1 - \beta_1x_i) - 0$ |
| $pr(y_i = 2)$ | $\Phi(\tau_2 - \beta_1 x_i) - \Phi(\tau_1 - \beta_1 x_i)$ |
| $pr(y_i = 3)$ | $1 - \Phi(\tau_2 - \beta_1 x_i)$ | 
]


<br>

Recall that this is a _probability distribution_, where the bounds fall between 0 and 1. Thus, why the lower bound is 0 and the upper bound is 1. 

---

### Predicted Probabilities

For example, say $\tau_1 = -1$, $\tau_2 = 1$, $\beta_1 =.05$. 

We can calculate the predicted probabilities of $y$ equalling a 1, 2 or 3 when $x = 15$ as follows

<br>

$$pr(y_i = 1) = \Phi((-1) - (.05)(15)) = .04$$

$$pr(y_i = 2) = \Phi((1) - (.05)(15)) - \Phi((-1) - (.05)(15)) = .56$$

$$pr(y_i = 3) = 1 - \Phi((1) - (.05)(15)) = .40$$

<br>

In this example, for an observation where $x = 15$, the probability that observed response will be 1 is 4%, 2 is 56%, and 3 is 40%. 

---

### Example: Student Support for the Iraq War

Survey data draw from a sample of 500 college students in 2002 leading up to the Iraq war. The survey asked whether college students agreed with the United States entering into the Iraq War. 

--

Three variables: 
  + `warsup`:  whether the respondent somewhat or strongly oppose/supported the War in Iraq
  
  + `dem`: 1 if the repondent was a democrat, 0 otherwise.
  
  + `female`: 1 if the repondent is female, 0 otherwise.

--

<br> 
Our hypothesis is that democratic respondents are less likely to support the Iraq War than non-democratic respondents.


---

```{r,echo=F}
dat <- read_csv('student_vote.csv')
dat$warsup = factor(dat$warsup,levels=c('strongly oppose',
                                        "somewhat oppose",
                                        "somewhat support",
                                        "strongly support"))
```

<br><br><br><br>
```{r}
summary(dat)
```

---

```{r}
war_ordered <- MASS::polr(warsup ~ dem + female, # Model
                          method = "probit", # Probit model
                          data = dat)
summary(war_ordered,digits = 2)
```

---

What is the probability of "Somewhat Supporting" the war given one is a democrat?

--

<br>

```{r}
# Grab the coefficients
B = war_ordered$coefficients
B
```

<br>

```{r}
# Grab the cut points
cuts = war_ordered$zeta
cuts
```

---

What is the probability of "Somewhat Supporting" the war given one is a democrat?

<br>

```{r}
# Grab the data
X = model.matrix(war_ordered)[,-1] # drop intercept
head(X)
```


---

What is the probability of "Somewhat Supporting" the war given one is a democrat?

<br>
```{r}
# Manipulate one value
X[,1] = 1 # Respondent is a democrat
```

<br>
Calculate the predicted probability for the "Somwhat support" category.

```{r}
pr_some_supp = pnorm(cuts[3] - X%*%B) - pnorm(cuts[2] - X%*%B)
mean(pr_some_supp)
```

---

What is the probability of "Somewhat Supporting" the war given one is _not_ a democrat?

<br>
```{r, highlight=T}
# Manipulate one value
X[,1] = 0 # Respondent is NOT a democrat #<<
```

<br>
Calculate the predicted probability for the "Somwhat support" category.

```{r}
pr_some_supp = pnorm(cuts[3] - X%*%B) - pnorm(cuts[2] - X%*%B)
mean(pr_some_supp)
```

---

<br><br>
The discrete difference between "Somewhat Supporting" the war given one...

<br>

- ...is a democrat = **17.7%**

- ...is _not_ a democrat = **35.6%**

<br><br>
The probability of "Somewhat Supporting" the war **decreases by 17.8%** if the respondent is a Democrat verse if he or she is not.

---

### Simulating Confidence Intervals

```{r,cache=T}
require(obsval) # Load the package

# Re-estimate the model using obsval
war_ordered2 <- 
  obsval(warsup ~ dem + female,
         data = dat,
         ci = .95,
         n.draws = 1000,
         reg.model = "oprobit",
         effect.var = "dem",
         effect.vals = c(0,1))
```

```{r}
# Effect summary
round(war_ordered2$effect_sum,3)
```

---

<br>
```{r}
war_ordered2$high.ci
```

```{r}
war_ordered2$means
```

```{r}
war_ordered2$low.ci
```


---

```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
# Function extracts the 2.5%, mean, and 97.5% interval
extract = function(x) c(quantile(x,.025),ave_pred=mean(x),quantile(x,.975))

# Apply to systematically draw it
non_dem = apply(war_ordered2$preds[,,1],2,extract)
dem = apply(war_ordered2$preds[,,2],2,extract)

non_dem = data.frame(t(non_dem)) %>% 
  rownames_to_column("category") %>% 
  mutate(democrat = 0)
dem = data.frame(t(dem)) %>% 
  rownames_to_column("category") %>% 
  mutate(democrat = 1)
predicted_effects_ord = bind_rows(dem,non_dem)

predicted_effects_ord %>%
  mutate(category = factor(category,levels = levels(dat$warsup))) %>% 
  ggplot(aes(category,ave_pred,color=factor(democrat))) + 
  geom_point(size=3) +
  geom_errorbar(aes(x=category,ymin=`X2.5.`,ymax=`X97.5.`),
                    width=.1,size=1) +
  labs(y="Predicted Probability",
       x="\nDo you think the U.S. should Invade Iraq?",
       color="") +
  scale_color_manual(values=c("darkred","steelblue"),
                     labels=c("Not Democrat","Democrat")) +
  theme_bw() +
  theme(legend.position = "top",
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 16),
        legend.text = element_text(family='serif',face='bold',size = 20)) 
```

---

class:newsection

# Multinomial Outcomes

---

### Multinomial Outcomes

- A nomial outcome variable is a variable with **discrete categories but no intrinsic ordering** 

- See this when a variable is just "names" or stated preferences. 

- The focus is analyzing **questions of "what", "which", "who", and "where"**, rather than questions regarding "how much" or "how many".

- Examples: 
  + Who did you vote for? 
  + Which product did you purchase?
  + What kind of job do you have?
  + What is the ethnicity of a new hire?
  
---

Say we have **three candidates** in an election: 

```{r,echo=F,fig.align="center",fig.width=9,fig.height=4.5,cache=T,dpi=300}
a <- 
  tibble(y=rep(3,3),
         x = c(1,3,5),
         candidate= c("Sally","Jack","Bruce"),
         size=c(75,75,75)) 
a %>% 
  ggplot(aes((x),(y),color=candidate,size=size)) +
  geom_point() +
  geom_text(color="white",aes(label=candidate),size=10) +
  scale_color_manual(values = c("steelblue","darkred","#4f1a70","grey30")) +
  scale_size_identity() +
  ylim(0,4) + xlim(0,6) +
  theme_void() +
  theme(legend.position = "none")
```


---

Say we have **three candidates** in an election: 

```{r,echo=F,fig.align="center",fig.width=9,fig.height=4.5,cache=T,dpi=300}
a <- 
  tibble(y=rep(3,3),
         x = c(1,3,5),
         candidate= c("Sally","Jack","Bruce"),
         size=c(75,75,75))  
b <- tibble(y = 1,
            x = 3,
            candidate= c("Voter"),
            size=50)
D = expand.grid(2,1:3)
bind_rows(a,b) %>% 
  ggplot(aes((x),(y),color=candidate,size=size)) +
  geom_segment(x=3,xend=5,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=3,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=1,y=1,yend=3,size=1,alpha=.5) +
  geom_point() +
  geom_text(color="white",aes(label=candidate),size=10) +
  scale_color_manual(values = c("steelblue","darkred","#4f1a70","grey30")) +
  scale_size_identity() +
  ylim(0,4) + xlim(0,6) +
  theme_void() +
  theme(legend.position = "none")
```

A voter is deciding which candidate to vote for. 

How do we figure out the probability of voting for one candidate (say, Sally) given one can also vote for Bruce or Jack?

---

One solution would be to treat the problem as a binary response.

```{r,echo=F,fig.align="center",fig.width=9,fig.height=4.5,cache=T,dpi=300}
a <- 
  tibble(y=rep(3,3),
         x = c(1,3,5),
         candidate= c("Sally","Jack","Bruce"),
         size=c(75,75,75))  
b <- tibble(y = 1,
            x = 3,
            candidate= c("Voter"),
            size=50)
D = expand.grid(2,1:3)
bind_rows(a,b) %>% 
  ggplot(aes((x),(y),color=candidate,size=size)) +
  geom_segment(x=3,xend=5,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=3,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=1,y=1,yend=3,size=1,alpha=.5) +
  geom_point() +
  geom_text(color="white",aes(label=candidate),size=10) +
  scale_color_manual(values = c("grey90","grey90","#4f1a70","grey30")) +
  scale_size_identity() +
  ylim(0,4) + xlim(0,6) +
  theme_void() +
  theme(legend.position = "none")
```

We just calculated the probability of voting for Sally (1) or not (0). 


**$$pr(vote_{sally} = 1) = \frac{exp(x_i\beta^{sally})}{1 + exp(x_i\beta^{sally})}$$**

---

One solution would be to treat the problem as a binary response.

```{r,echo=F,fig.align="center",fig.width=9,fig.height=4.5,cache=T,dpi=300}
a <- 
  tibble(y=rep(3,3),
         x = c(1,3,5),
         candidate= c("Sally","Jack","Bruce"),
         size=c(75,75,75))  
b <- tibble(y = 1,
            x = 3,
            candidate= c("Voter"),
            size=50)
D = expand.grid(2,1:3)
bind_rows(a,b) %>% 
  ggplot(aes((x),(y),color=candidate,size=size)) +
  geom_segment(x=3,xend=5,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=3,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=1,y=1,yend=3,size=1,alpha=.5) +
  geom_point() +
  geom_text(color="white",aes(label=candidate),size=10) +
  scale_color_manual(values = c("steelblue","grey90","grey90","grey30")) +
  scale_size_identity() +
  ylim(0,4) + xlim(0,6) +
  theme_void() +
  theme(legend.position = "none")
```

We can just re-arrange to calculate the probability another candidate. 
**$$pr(vote_{bruce} = 1) = \frac{exp(x_i\beta^{bruce})}{1 + exp(x_i\beta^{bruce})}$$**

---

One solution would be to treat the problem as a binary response.

```{r,echo=F,fig.align="center",fig.width=9,fig.height=4.5,cache=T,dpi=300}
a <- 
  tibble(y=rep(3,3),
         x = c(1,3,5),
         candidate= c("Sally","Jack","Bruce"),
         size=c(75,75,75))  
b <- tibble(y = 1,
            x = 3,
            candidate= c("Voter"),
            size=50)
D = expand.grid(2,1:3)
bind_rows(a,b) %>% 
  ggplot(aes((x),(y),color=candidate,size=size)) +
  geom_segment(x=3,xend=5,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=3,y=1,yend=3,size=1,alpha=.5) +
  geom_segment(x=3,xend=1,y=1,yend=3,size=1,alpha=.5) +
  geom_point() +
  geom_text(color="white",aes(label=candidate),size=10) +
  scale_color_manual(values = c("grey90","darkred","grey90","grey30")) +
  scale_size_identity() +
  ylim(0,4) + xlim(0,6) +
  theme_void() +
  theme(legend.position = "none")
```

We can just re-arrange to calculate the probability another candidate. 
**$$pr(vote_{jack} = 1) = \frac{exp(x_i\beta^{jack})}{1 + exp(x_i\beta^{jack})}$$**

---

### Multinomial model

The multinomial logit model allows us to estimate all these binary logits for the different outcome categories simultaneously.

<br>

We can estimate the probability of making one choice _given all the alternative options_.

<br>

$$pr(vote = sally) = \frac{exp(x_i\beta^{sally})}{exp(x_i\beta^{bruce}) + exp(x_i\beta^{jack}) + exp(x_i\beta^{sally})}$$

---

### Identification

To identify the model, we need to constrain the coefficients for one of the choices (i.e. candidates) to equal 0. This choice will be the **reference category**.

From our candidates, let's choose Bruce to be the reference category: $$\beta^{bruce}=0$$ 

Thus, the multinomial logit model is as follows:

<br>

$$pr(vote = sally) = \frac{exp(x_i\beta^{sally})}{1 + exp(x_i\beta^{jack}) + exp(x_i\beta^{sally})}$$

---

### Predicted Probability

To figure out probability of voting for any of the other candidates, we just need to re-arrange this equation.

<br>

$$pr(vote = jack) = \frac{exp(x_i\beta^{jack})}{1 + exp(x_i\beta^{jack}) + exp(x_i\beta^{sally})}$$

<br><br>

$$pr(vote = bruce) = \frac{1}{1 + exp(x_i\beta^{jack}) + exp(x_i\beta^{sally})}$$

---

### Predicted Probability

Say $x_i$ is a dummy variable that takes on the value of 0 and 1. To calculate the probability of voting for **Sally** given $x_i=1$ we do as follow:

$$\frac{exp(\beta_0^{sally} + \beta_1^{sally} x_i)}{1 + exp(\beta_0^{jack} + \beta_1^{jack} x_i) + exp(\beta_0^{sally} + \beta_1^{sally} x_i)}$$

<br>

** $$\frac{exp(\beta_0^{sally} + \beta_1^{sally} (1))}{1 + exp(\beta_0^{jack} + \beta_1^{jack} (1)) + exp(\beta_0^{sally} + \beta_1^{sally} (1))}$$ **

---

### Predicted Probability

Say $x_i$ is a dummy variable that takes on the value of 0 and 1. To calculate the probability of voting for **Jack** given $x_i=1$ we do as follow:

$$\frac{exp(\beta_0^{jack} + \beta_1^{jack} x_i)}{1 + exp(\beta_0^{jack} + \beta_1^{jack} x_i) + exp(\beta_0^{sally} + \beta_1^{sally} x_i)}$$

<br>

** $$\frac{exp(\beta_0^{jack} + \beta_1^{jack} (1))}{1 + exp(\beta_0^{jack} + \beta_1^{jack} (1)) + exp(\beta_0^{sally} + \beta_1^{sally} (1))}$$ **

---

### Predicted Probability

Say $x_i$ is a dummy variable that takes on the value of 0 and 1. To calculate the probability of voting for **Bruce** given $x_i=1$ we do as follow:

$$\frac{1}{1 + exp(\beta_0^{jack} + \beta_1^{jack} x_i) + exp(\beta_0^{sally} + \beta_1^{sally} x_i)}$$

<br>

** $$\frac{1}{1 + exp(\beta_0^{jack} + \beta_1^{jack} (1)) + exp(\beta_0^{sally} + \beta_1^{sally} (1))}$$ **

<br>

and so on...

---

### Discrete Differences in the Predicted Probability

<br><br>
If we wanted to calculate the discrete difference for voting for Sally given different values of $x$, we can simply calculate the difference in probabilities as we did with other models. 

<br><br>

$$pr(vote = sally | x = 1) - pr(vote = sally | x = 0)$$

---

### IIA Assumption 

<br>

- A key assumption for a multinomial choice model is the **independence of irrelevant alternatives assumption (IIA)**. 

- The idea is that the relative probability of existing alternatives (choices) is not affected by changes to the choice set (i.e. adding or removing choices). 

- Adding and dropping a choice doesn't impact the underlying probability of the choices left on the table. 

- If this assumption is violated, estimates will be biased.

---

Say we had commuters deciding between three modes of travel: bus, train, and car.

Each segment represents the probability of choosing that transportation option. Here the probability of choosing any one mode of transportation is roughly equal.

```{r,echo=F,fig.align="center",fig.width=10,fig.height=5,cache=T,dpi=300}
tibble(options = factor(c("car",'train',"bus"),levels=c("bus","car",'train')),
       pos = rep(1,3),
       probs = c(.32,.34,.34)) %>% 
  ggplot(aes(pos,probs,fill=options)) +
  geom_bar(position = "stack",stat = "identity") +
  coord_flip() + 
  labs(fill="",y="\nProbabilities") +
  theme_minimal() +
  scale_fill_manual(values = c("black","grey30","grey70")) +
  theme(legend.position = "top",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text.x = element_text(family='serif',face='bold',size = 20),
        legend.text = element_text(family='serif',face='bold',size = 25))
```

---

Say we had commuters deciding between three modes of travel: bus, train, and car.

Each segment represents the probability of choosing that transportation option. Here it is not.
<br><br>

```{r,echo=F,fig.align="center",fig.width=10,fig.height=5,cache=T,dpi=300}
tibble(options = factor(c("car",'train',"bus"),levels=c("bus","car",'train')),
       pos = rep(1,3),
       probs = c(.5,.33,.17)) %>% 
  ggplot(aes(pos,probs,fill=options)) +
  geom_bar(position = "stack",stat = "identity") +
  coord_flip() + 
  labs(fill="",y="\nProbabilities") +
  theme_minimal() +
  scale_fill_manual(values = c("black","grey30","grey70")) +
  theme(legend.position = "top",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text.x = element_text(family='serif',face='bold',size = 20),
        legend.text = element_text(family='serif',face='bold',size = 25))
```

---

Say we had commuters deciding between three modes of travel: bus, train, and car.

Now say that we added another option to the table: the chance to ride a <font color ="red">red</font> or <font color ="blue">blue</font> bus. As long as the addition of these new choices doesn't change the probability of choosing any other choices, IIA is satisfied.
<br>

```{r,echo=F,fig.align="center",fig.width=10,fig.height=5,cache=T,dpi=300}
tibble(options = factor(c("car",'train',"red bus","blue bus"),
                        levels=c("red bus","blue bus","car",'train')),
       pos = rep(1,4),
       probs = c(.5,.33,.085,.085)) %>% 
  ggplot(aes(pos,probs,fill=options)) +
  geom_bar(position = "stack",stat = "identity") +
  coord_flip() + 
  labs(fill="",y="\nProbabilities") +
  theme_minimal() +
  scale_fill_manual(values = c("red","blue","grey30","grey70")) +
  theme(legend.position = "top",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text.x = element_text(family='serif',face='bold',size = 20),
        legend.text = element_text(family='serif',face='bold',size = 25))
```

---

Say we had commuters deciding between three modes of travel: bus, train, and car.

Now say that we added another option to the table: the chance to ride a <font color ="red">red</font> or <font color ="blue">blue</font> bus. If this is not the case (new bus colors made you more likely to take the bus over the train or car), then <font color = "darkred"> IIA would be violated </font>. 

```{r,echo=F,fig.align="center",fig.width=10,fig.height=5,cache=T,dpi=300}
tibble(options = factor(c("car",'train',"red bus","blue bus"),
                        levels=c("red bus","blue bus","car",'train')),
       pos = rep(1,4),
       probs = c(.4,.23,.085,.285)) %>% 
  ggplot(aes(pos,probs,fill=options)) +
  geom_bar(position = "stack",stat = "identity") +
  coord_flip() + 
  labs(fill="",y="\nProbabilities") +
  theme_minimal() +
  scale_fill_manual(values = c("red","blue","grey30","grey70")) +
  theme(legend.position = "top",
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        panel.grid.major = element_blank(),
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text.x = element_text(family='serif',face='bold',size = 20),
        legend.text = element_text(family='serif',face='bold',size = 25))
```

---

### Example: Student Support for the Iraq War

Let's use the survey data from of 500 college students in 2002 leading up to the Iraq war. 

```{r,echo=F}
dat <- read_csv('student_vote.csv')
dat$warsup = factor(dat$warsup,levels=c('strongly oppose',
                                        "somewhat oppose",
                                        "somewhat support",
                                        "strongly support"))
```
<br>
```{r}
summary(dat)
```
<br>
> This time rather than treat the outcome as an ordered categories, we'll treat it as **discrete choices**.

---
```{r}
multi_mod <- nnet::multinom(warsup ~ dem + female,data=dat)
multi_mod
```

---

<br><br>
```{r,highlight=T}
broom::tidy(multi_mod, 
            exponentiate = F #<<
            ) %>% 
  mutate_if(is.numeric, function(x) round(x,3))
```

---

What is the probability of "Somewhat Supporting" the war given one is a democrat?

--

```{r}
# Extract the coefficients

B_all <- coefficients(multi_mod)
B_all
```
<br>
```{r}
B_so = B_all[1,] # Coefs for "somewhat oppose"
B_ss = B_all[2,] # Coefs for "somewhat support"
B_SS = B_all[3,] # Coefs for "strongly support"
```

---

What is the probability of "Somewhat Supporting" the war given one is a democrat?

```{r}
# Extract the data 

X = model.matrix(multi_mod)
X
```


---

What is the probability of "Somewhat Supporting" the war given one is a democrat?

```{r}
# Manipulate one value 
X[,2] = 1 # Respondent is a democrat
```

Calculate the predicted probability for the "Somwhat support" choice category. 
```{r}
# Denominator of all options
denominator <- 1 + exp(X%*%B_so) + exp(X%*%B_ss) + exp(X%*%B_SS)

# Probability of "Somewhat Support"
pr_some_supp = exp(X%*%B_ss)/denominator

# Average across all observations
mean(pr_some_supp)
```

---

What is the probability of "Somewhat Supporting" the war given one is _not_ a democrat?

```{r,highlight=T}
# Manipulate one value 
X[,2] = 0 # Respondent is not a democrat #<<
```

Calculate the predicted probability for the "Somwhat support" choice category. 
```{r}
# Denominator of all options
denominator <- 1 + exp(X%*%B_so) + exp(X%*%B_ss) + exp(X%*%B_SS)

# Probability of "Somewhat Support"
pr_some_supp = exp(X%*%B_ss)/denominator

# Average across all observations
mean(pr_some_supp)
```

---

<br><br>
The discrete difference between "Somewhat Supporting" the war given one...

<br>

- ...is a democrat = **14.4%**

- ...is _not_ a democrat = **38.2%**

<br><br>
The probability of "Somewhat Supporting" the war **decreases by 23.8%** if the respondent is a Democrat verse if he or she is not.

---

### Simulating Confidence Intervals

```{r,highlight=T,cache=T}
require(obsval) # Load the package

# Re-estimate the model using obsval
multi_mod2 <- 
  obsval(warsup ~ dem + female,
         data = dat,
         ci = .95,
         n.draws = 1000,
         baseline.category = "strongly oppose", #<<
         reg.model = "mlogit",
         effect.var = "dem",
         effect.vals = c(0,1))
```

---

### Simulating Confidence Intervals

```{r}
round(multi_mod2$effect_sum,3)
```

---

<br>
```{r}
multi_mod2$high.ci
```

```{r}
multi_mod2$means
```

```{r}
multi_mod2$low.ci
```

---

```{r,echo=F,fig.align="center",fig.width=10,fig.height=8,cache=T,dpi=300}
# Function extracts the 2.5%, mean, and 97.5% interval
extract = function(x) c(quantile(x,.025),ave_pred=mean(x),quantile(x,.975))

categories = c("Strongly Oppose","Somewhat Oppose",
               "Somewhat Support","Strongly Support")

# Apply to systematically draw it
non_dem = apply(multi_mod2$preds[,,1],2,extract)
dem = apply(multi_mod2$preds[,,2],2,extract)

non_dem = data.frame(t(non_dem)) %>% 
  rownames_to_column("category") %>% 
  mutate(democrat = 0)
dem = data.frame(t(dem)) %>% 
  rownames_to_column("category") %>% 
  mutate(democrat = 1)
predicted_effects_nom = bind_rows(dem,non_dem)

predicted_effects_nom %>%
  mutate(category = factor(category,levels = tolower(categories))) %>% 
  ggplot(aes(category,ave_pred,color=factor(democrat))) + 
  geom_point(size=3) +
  geom_errorbar(aes(x=category,ymin=`X2.5.`,ymax=`X97.5.`),
                    width=.1,size=1) +
  labs(y="Predicted Probability",
       x="\nDo you think the U.S. should Invade Iraq?",
       color="") +
  scale_color_manual(values=c("darkred","steelblue"),
                     labels=c("Not Democrat","Democrat")) +
  theme_bw() +
  theme(legend.position = "top",
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 16),
        legend.text = element_text(family='serif',face='bold',size = 20)) 
```


---

### Compare

.pull-left[
.center[**Ordered**]
```{r,echo=F,fig.align="center",fig.height=9,fig.width=8,cache=T,dpi=300}
predicted_effects_ord %>%
  mutate(category = factor(category,levels = levels(dat$warsup))) %>% 
  ggplot(aes(category,ave_pred,color=factor(democrat))) + 
  geom_point(size=5) +
  geom_errorbar(aes(x=category,ymin=`X2.5.`,ymax=`X97.5.`),
                    width=.1,size=2) +
  labs(y="Predicted Probability",
       x="\nDo you think the U.S. should Invade Iraq?",
       color="") +
  scale_color_manual(values=c("darkred","steelblue"),
                     labels=c("Not Democrat","Democrat")) +
  theme_bw() +
  ylim(0,.65) +
  theme(legend.position = "top",
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 16),
        legend.text = element_text(family='serif',face='bold',size = 20)) 
```
]

.pull-right[
.center[**Nominal**]
```{r,echo=F,fig.align="center",fig.height=9,fig.width=8,cache=T,dpi=300}
predicted_effects_nom %>%
  mutate(category = factor(category,levels = tolower(categories))) %>% 
  ggplot(aes(category,ave_pred,color=factor(democrat))) + 
  geom_point(size=5) +
  geom_errorbar(aes(x=category,ymin=`X2.5.`,ymax=`X97.5.`),
                    width=.1,size=2) +
  labs(y="",
       x="\nDo you think the U.S. should Invade Iraq?",
       color="") +
  scale_color_manual(values=c("darkred","steelblue"),
                     labels=c("Not Democrat","Democrat")) +
  ylim(0,.65) +
  theme_bw() +
  theme(legend.position = "top",
        plot.title = element_text(family='serif',face='bold',size = 20),
        axis.title = element_text(family='serif',face='bold',size = 20),
        axis.text = element_text(family='serif',face='bold',size = 16),
        legend.text = element_text(family='serif',face='bold',size = 20)) 
```
]
